---
title: "Time Series Lab"
author:
- "Michael McCormack"
- "Noah Johnson"
date: "March 15, 2018"
output: html_document
---

```{r, include = FALSE}
library(install.load)
install_load('tidyverse')
install_load('reshape2')
install_load('ggfortify')
install_load('forecast')
```

## Step 1: Get the data.  
Download from [website](https://cdn.rawgit.com/mikejt33/DataViz/246c2026/data/flights.csv.gz).
```{r}
flights <- read.csv('flights.csv.gz')
```

## Step 2: Prepare the data.  

```{r}
str(flights)
unique(flights$X)
```

The X column is just NAs, so we can remove it.

```{r}
flights <- flights %>% select(-c(X))
```

*Are there any null values?

```{r}
sapply(flights, function(y) sum(is.na(y)))
```

If departure or arrival delay is NA, does that mean there was just 0 delay?

```{r}
head(flights %>% filter(is.na(ARR_DELAY)))
```

It looks like there are rows that have a flight date, and airline carrier, but NAs for delays and departure time. I'm not confident we can assume this mean there was no delay. Maybe it just means the data enterer knew the date of the flight but not how long it took, or if there were delays. So rather than converting these NAs to zeroes, let's just drop these rows.

```{r}
flights <- flights %>% drop_na()
```

Time series data needs to be over a regular time interval. Calculate the average departure delay time and/or average arrival delay time for each day of 2017.

```{r}
flights %>% group_by(FL_DATE) %>% summarise(avgDepDelay=mean(DEP_DELAY), avgArrivDelay=mean(ARR_DELAY))
```

If you like, compare average delay times for different carriers or different airports by creating multiple time series.

Let's look at Spirit Airlines (cheap) vs Delta Airlines (expensive)
```{r}
spirit.delays <- flights %>% filter(CARRIER=='NK') %>% group_by(FL_DATE) %>% summarise(avgDepDelay=mean(DEP_DELAY), avgArrivDelay=mean(ARR_DELAY))
as.ts(spirit.delays)

delta.delays <- flights %>% filter(CARRIER=='NK') %>% group_by(FL_DATE) %>% summarise(avgDepDelay=mean(DEP_DELAY), avgArrivDelay=mean(ARR_DELAY))
```

## Step 3: Create a ts object of the data.  
Refer to the slides for tips on how to do this.
```{r}

```

## Step 4: Plot the time series using base package and ggplot (advanced).
Create a basic visualization of the time series.

```{r}

```

## Step 5: Smooth the data to reduce noise and identify trends.

### Create your own simple moving average for monthly data. Plot the smoothed data using base package. Plot both the originial and the smoothed data ggplot (advanced). 

__Hints__
* good StackOverflow reference for moving average in R: https://stackoverflow.com/questions/743812/calculating-moving-average
* watch out for functions that may have been masked by other packages
* ggplot: may need to convert data to long format to plot mutliple series  

```{r}

```

__Questions__  
1. How does the neighborhood size, i.e. the number of points in each localized subset, affect the amount of smoothing?  
2. What happened to endpoints of the smoothed data?  

### Advanced: Smooth the same data using Local Regression (loess). Plot smoothed data using base package. Plot all three series (original, smoothed by MA, and smoothed by loess) using ggplot (advanced).

__Hint__  
* loess() requires all predictors to be numerical so dates cannot be used  

Try different values for the span argument and see how it affects the amount of smoothing.  

```{r}

```

# Dive in Deeper to TimeSeries  

For this portion of our lab we will be using data from the AirPassengers Dataset  

```{r}
data(AirPassengers)
```

## Step 6: Make an inital TimeSeries Visual of the data  
```{r}

```
## Step 7: Compute the Moving Average of this data using forecast package and vizualize this  
```{r}

```
## Step 8: Remove the Trend from the data and Visualize this  
```{r}

```
## Step 9: Create a decomposition of the data by month  
-- Hint (Frequency = 12)
```{r}

```

